%!TEX root = ../thesis.tex
\ifpdf
    \graphicspath{{Sections/3_methods_figs/Raster/}{Sections/3_methods_figs/PDF/}{Sections/3_methods_figs/}}
\else
    \graphicspath{{Sections/3_methods_figs/Vector/}{Sections/3_methods_figs/}}
\fi

\chapter{Methods}
\label{sec:method}

\section{The Dataset}
In this work, we use the LIDC-IDRI dataset \cite{armato_iii_data_2015-1,armato_lung_2011}, which contains 1018 chest CT scans from 1010 patients distributed as DICOM images. We retain one scan per patient, and sorted patient-IDs are assigned to train, validation and test sets, meaning that no patient contributes slices to more than one set. The default PaDIS regime selects a total of at most $4$ slices per patient, capped by a requirement for equal numbers of nondule-containing and non-nodule-containing slices. The detailed breakdown is presented in \autoref{tab:lidc_split_sizes}.

\begin{table}[ht] 
    \centering 
    \begin{tabular}{lccc} 
        \hline \textbf{Split} & \textbf{Patient-ID slots} & \textbf{Images in 4-slice regime} & \textbf{Images in all-slice regime} \\
        \hline 
        Train & 809 & 2713 & 189725 \\ 
        Validation & 101 & 328 & 27426 \\ 
        Test & 102 & 326 & 25719 \\ 
        \hline Total & 1012 & 3367 & 242870 \\ 
        \hline 
    \end{tabular}
    \caption[Processed LIDC-IDRI split sizes.]{Processed LIDC-IDRI split used for the completed experiments. The 1012 sorted patient-IDs include two with no data available.}
    \label{tab:lidc_split_sizes} 
\end{table}

\section{Experimental Geometry}
In this work, we use a 2D fan-beam CT geometry implemented in LION \cite{kiss_benchmarking_2025,noauthor_cambridgecialion_2026}. LION's operator stack uses tomosipo \cite{hendriksen_tomosipo_2021} over the ASTRA Toolbox \cite{van_aarle_fast_2016}, while the analytic reconstruction helper is provided by \texttt{ts\_algorithms} \cite{hendriksen_ahendriksents_algorithms_2026}. Each LIDC-IDRI slice is treated independently with a $300\,\mathrm{mm}\times300\,\mathrm{mm}$ in-plane field of view. The default experiments reconstruct a $256\times 256$ grid, whilst the explicitly labelled native-resolution experiments use a $512\times 512$ grid. The detector contains $900$ bins over a $900\, \mathrm{mm}$ detector width, with a $575\,\mathrm{mm}$ source-to-origin distance and a $1050\,\mathrm{mm}$ source-to-detector distance. The simulated measurements are noise-free projections of normalised intensity rather than calibrated attenuation, isolating sparse sampling and limited angular coverage from photon-count statistics.

We use fan-beam geometry to test whether the PaDIS conclusions transfer beyond Hu et al.'s parallel-beam setup and to reproduce the finite-source divergence used by clinical scanners \cite{wunderlich_image_2008}. This choice, however, still does not make the experiment clinically realistic because the operator remains two-dimensional, noise-free, and omits scatter, beam hardening, motion and cone-beam effects.

\section{Model Training}
\label{sec:methods-model-training}

\subsection{Diffusion Model Training}
Diffusion-prior training uses the variance-exploding regime in \secref{sec:score_based_sdes}, so the denoisers provide scores for unconditional generation and inverse problems as explained in \secref{subsec:diffusion_models_inverse_problems} and \secref{subsec:padis_approach}.

Diffusion denoisers use the noise-conditional NCSN++ model, $F_{\boldsymbol{\phi}}$ \cite{song_score-based_2020-1}, with EDM preconditioning from \secref{subsec:edm_preconditioning}. Noise is passed to every residual block, and default inputs comprise the noisy image and $[-1,1]$ normalised pixel coordinates, while the position ablation receives only the image. Each model has four resolution levels, four residual blocks per level, 128 base channels, dropout 0.05 and attention at resolution 16. Patch channel counts are $128$, $256$, $256$ and $256$, while whole-image counts are $128$, $256$, $256$ and $512$.

All diffusion models used Adam \cite{kingma_adam_2017-1} with learning rate $2\times10^{-4}$, ramped over the first 10 million examples or patches. EDM used $\sigma_{\mathrm{data}}=0.5$, loss weight $\lambda(\sigma)=(\sigma^2+\sigma_{\mathrm{data}}^2)/(\sigma\sigma_{\mathrm{data}})^2$, and $\log\sigma\sim\mathcal{N}(-1.2,1.2^2)$ truncated to $[0.002,40]$ \cite{karras_elucidating_2022-1}. On an RTX PRO 6000, whole-image models trained for 24 hours and patch-based models for 12. Their sample counts are listed in \hyperref[app:training-exposure]{Appendix~\ref*{app:training-exposure}}.

Whole-image denoisers are trained on complete $256\times256$ slices. The default patch model samples sizes 56, 32 and 16 with probabilities 0.5, 0.3 and 0.2, while the native resolution model replaces 56 by 64. Ablations investigate using much more training data (all processed slices), removing position channels, and alternative patch sizes with training schedules listed in \autoref{tab:patch_size_schedules}.

Validation was performed throughout training. However, during the final six hours, it was conducted more frequently and over more validation samples. At each validation, the mean EDM loss was evaluated using an exponential moving average (EMA) of the model parameters. The EMA parameter state with the lowest loss during this phase was saved as \texttt{min\_intense\_val} and used for reconstruction and generation.

\begin{table}[ht] 
    \centering 
    \begin{tabular}{lccc} 
        \hline 
        \textbf{Ablation} & \textbf{Sampled patch sizes} & \textbf{Sampling probabilities} & \textbf{Padding width} \\ 
        \hline 
        $P=8$ & $8$ & $1.0$ & $8$ \\ 
        $P=16$ & $8,\ 16$ & $0.3,\ 0.7$ & $16$ \\ 
        $P=32$ & $8,\ 16,\ 32$ & $0.2,\ 0.3,\ 0.5$ & $32$ \\ 
        $P=56$ & $16,\ 32,\ 56$ & $0.2,\ 0.3,\ 0.5$ & $24$ \\ 
        $P=96$ & $32,\ 64,\ 96$ & $0.2,\ 0.3,\ 0.5$ & $32$ \\ 
        \hline 
    \end{tabular} 
    \caption[PaDIS patch-size training schedules.]{Patch-size schedules used for PaDIS patch-size ablations. Each training step samples one patch size according to the listed probabilities. The $P=56$ row is the default.}
    \label{tab:patch_size_schedules}
\end{table}

\subsection{PnP Denoiser Training}
PnP training uses two DRUNet denoisers \cite{zhang_plug-and-play_2022} trained separately for LION's PnP-ADMM reconstruction \cite{chan_plug-and-play_2017}. Each training step samples Gaussian noise uniformly from $\sigma \in[0,0.05]$ and adds it to the clean image, from which the model learns to recover the clean target.

Both models use the four-slice split, Adam at $10^{-4}$ for 100 epochs, batch size 8 and MSE loss. One receives only the noisy image, while the other adds noise level as a constant channel. Both have 64 internal channels, four blocks per level, bias-free convolutions and leaky-ReLU activations. We use the lowest-validation-loss checkpoint from the epoch-wise checks, supplying the conditioned model with noise level 0.03 during reconstruction.

\section{Reconstructor Implementation}
\label{sec:methods-reconstructor-implementation}

The original PaDIS paper and released code specify some methods differently (\secref{sec:discussion-reproducibility}) \cite{hu_learning_2024,hu_jasonhu4padis_2026-1}, so we choose to re-implemented both versions to compare. The Paper-Faithful implementationfollows the written algorithms, Repo-Faithful follows the released code, and LION-Physical is our operator-normalised extension. The reproduction tracks exist only where the original PaDIS work gives enough detail, whereas LION-Physical covers all methods. \autoref{tab:methods_by_implementation} summarises these components, with differences reported in \secref{subsec:repo-faithful-implementation} and \secref{subsec:lion-physical-implementations}.

\begin{table}[ht]
  \centering
  \caption[Reconstruction components by implementation track.]{Reconstruction components implemented in each track. A star indicates an additional modification or numerical stabilisation described below.}
  \label{tab:methods_by_implementation}
  \begin{tabular}{lccc}
  \hline
  \textbf{Component} & \textbf{Paper-Faithful} & \textbf{Repo-Faithful} & \textbf{LION-Physical} \\
  \hline
  Baseline FDK              & --             & --             & $\checkmark$ \\
  Chambolle--Pock TV        & --             & --             & $\checkmark^*$ \\
  PnP-ADMM                  & --             & --             & $\checkmark$ \\
  Whole-image prior         & $\checkmark$ & --             & $\checkmark$ \\
  PaDIS prior               & $\checkmark$ & $\checkmark$ & $\checkmark$ \\
  Fixed patch averaging     & --             & $\checkmark^*$ & $\checkmark$ \\
  Fixed patch stitching     & --             & $\checkmark^*$ & $\checkmark$ \\
  VE-DPS                    & $\checkmark$ & $\checkmark^*$ & $\checkmark$ \\
  Langevin                  & $\checkmark$ & $\checkmark^*$ & $\checkmark$ \\
  Predictor-corrector       & $\checkmark^*$ & $\checkmark^*$ & $\checkmark$ \\
  VE-DDNM                   & $\checkmark^*$ & $\checkmark^*$ & $\checkmark^*$ \\
  \hline
  \end{tabular}
\end{table}

Throughout, an outer step changes noise level, while an inner step repeats the sampler update at that level. Unless stated otherwise, diffusion reconstructions use 100 geometrically spaced outer levels from $\sigma_{\max}=10$ to $\sigma_{\min}=0.002$, with 10 inner updates each, totalling 1000 model evaluations. The 8-view experiments instead use $\sigma_{\min}=0.003$. The schedule ablation changes only the spacing of the outer levels from geometric to EDM-style. Predictor-corrector instead makes one predictor and one corrector evaluation for each of 99 adjacent pairs, totalling 198. Paper-Faithful and LION-Physical VE-DDNM use the one-evaluation-per-level layout of Algorithm A.3, with 1000 outer levels. Repo-Faithful VE-DDNM retains the released CT configuration of 100 levels with 10 inner updates.

For the analytic baseline, we apply LION's Hann-filtered FDK with frequency scaling 0.3 \cite{feldkamp_practical_1984}. For an iterative baseline, Hu et al. used ADMM-TV \cite{hu_learning_2024}. Here, we retain their isotropic TV objective \cite{rudin_nonlinear_1992} but use LION's CP implementation from \secref{subsec:iterative_reconstruction_denoising_priors} \cite{chambolle_first-order_2011}. For this method, we use $\lambda=0.001$, 1000 iterations, and output clipping in $[0,1]$. In-line with the LION defaults, our PnP-ADMM implementation, uses $\eta=3\times10^{-5}$, 60 outer iterations, at most 50 conjugate-gradient iterations, tolerance $10^{-7}$, and state output clipping in $[0,1]$.

The data updates within the VE-DPS sampler are varied across implementations. Where $F$ maps a normalised image to its sinogram, and $L_F=\lVert F\rVert_2^2$ can be estimated using 20 power iterations with tolerance $10^{-4}$ \cite{golub_matrix_2012,beck_fast_2009}, and we define $\mathbf{r}^k=\mathbf{y}-F\hat{\mathbf{x}}_0^k$, the implemented updates are

\begin{equation}
\mathbf{z}_{\mathrm{data}}^k=
\begin{cases}
\displaystyle \mathbf{z}^k-\frac{\zeta}{\lVert\mathbf{r}^k\rVert_2}
\nabla_{\mathbf{z}^k}\lVert\mathbf{r}^k\rVert_2^2,
& \text{Paper-Faithful},\\[6pt]
\displaystyle \mathbf{z}^k-\zeta c_{\mathrm{DPS}}
\nabla_{\mathbf{z}^k}\lVert\mathbf{r}^k\rVert_2,
& \text{Repo-Faithful},\\[6pt]
\displaystyle \mathbf{z}^k-\frac{\zeta}{L_F}
\nabla_{\mathbf{z}^k}\frac{1}{2}\lVert\mathbf{r}^k\rVert_2^2,
& \text{LION-Physical}.
\end{cases}
\label{eq:implementation_data_updates}
\end{equation}

Here, each gradient is taken through the diffusion denoiser. In Repo-Faithful, $c_{\mathrm{DPS}}=0.0405$ scales this VE-DPS gradient and $c_{\mathrm{adj}}=0.1022$ scales the direct $F^T\mathbf{r}^k$ correction in Langevin and both predictor-corrector updates.

Patch averaging and stitching use fixed layouts with an eight-pixel overlap.
Averaging takes the mean of all predictions covering each pixel, whilst
stitching allows later patches to replace earlier values in overlaps. Both use
VE-DPS sampling, but unlike the PaDIS assembly they do not repeatedly redraw shifted patch
partitions. These methods are motivated by the works of Ozdenizci et al.
\cite{ozdenizci_restoring_2023} and Rumberger et al. \cite{rumberger_how_2021},
but mimic the code implementations by Hu et al. \cite{hu_jasonhu4padis_2026-1}
rather than the original works.

Every VE-DDNM track clips its FDK approximate-inverse terms to prevent non-finite values, but Paper-Faithful and LION-Physical implementations also clip the corrected state. In the Langevin step, these two tracks use $\alpha_k=0.1\sigma_k^2$, while Repo-Faithful uses $\alpha_k=0.2\sigma_k^2$.

\subsection{Repo-Faithful Implementations}
\label{subsec:repo-faithful-implementation}
We implemented these methods to match the publicly available codebase, which differs from the descriptions in the original PaDIS paper in several important ways \cite{hu_learning_2024,hu_jasonhu4padis_2026-1}. Its VE-DPS implementation differentiates the norm of the residual and begins from FDK, whilst predictor-corrector evaluates the denoiser at $\sigma_k$, divides the score by $\sigma_{k+1}^2$ and reuses the predictor patch layout.

The Repo-Faithful track implements these rules, clipping and patch randomisation with fixed scales for LION geometry. Final experiments use the geometric schedule from the original PaDIS paper, with the public EDM-style schedule as an ablation. Translation validation used an adapted version of the released code in the companion \texttt{PaDIS-Lion-Compat} directory with the same slices, checkpoint and geometry. Across three matched PaDIS-VE-DPS reconstructions, we achieved cross-implementation SSIM of $0.996$, demonstrating close equivalence between the two codebases despite different implementation regimes.

\subsection{LION-Physical Implementations}
\label{subsec:lion-physical-implementations}
Our LION-Physical extension replaces fixed data-consistency scales with operator-derived scaling. The factor $L_F^{-1}$ also scales the direct $F^T\mathbf{r}$ corrections in Langevin and both predictor-corrector updates. It does not modify the diffusion prior or noise updates, the schedule, or the classical baselines. VE-DDNM bypasses these gradient and adjoint updates, so this factor is not applied to its FDK approximate-inverse correction. This makes the data-consistency coefficient more comparable between geometries, but does not guarantee sampler convergence.

Predictor-corrector follows the Paper-Faithful convention, with the corrector evaluated at the next level using a newly sampled patch layout. Within the otherwise matched Paper-Faithful VE-DDNM update, the selected LION-Physical patch configuration uses $\gamma_{\mathrm{noise}}=0.5$ rather than $1$, halving only the stochastic Langevin term.

\subsection{Hyperparameter Tuning}
Hyperparameter selection used only validation data. Finite runs were ranked by mean PSNR, treating differences below $0.1\,\mathrm{dB}$ as ties, then by SSIM within $0.01$, and finally by MAE. Where possible, one setting covered 8 and 20 views. Higher-view experiments use this setting without dedicated tuning, whilst the 512-resolution experiment was tuned separately. The \href{https://github.com/THartigan/MPhil-Project/blob/main/LION/scripts/paper_scripts/PaDIS-Reproduction/tuning/TUNING.md\#attempted-values}{sweep ranges} and \href{https://github.com/THartigan/MPhil-Project/blob/main/LION/scripts/paper_scripts/PaDIS-Reproduction/README.md\#tuned-reconstruction-defaults}{selected hyperparameters} are documented in LION. We did not tune the non-default patch-size and no-position ablations separately, so they inherit the default $P=56$ setting.

\section{Experiments}

The primary comparison uses noise-free sinograms from 8 and 20 uniformly spaced views over $360^\circ$ at $256\times256$. This comparison uses the components in \autoref{tab:methods_by_implementation}, which produce the results in \autoref{tab:main-quality}. Paper-Faithful and Repo-Faithful test the central PaDIS claim under two readings of the combined specification, while LION-Physical evaluates our extension on the same images, priors and measurements.

Selected methods are also evaluated with 60 full-angle views and with 20 views over a $120^\circ$ angular range, and is designed to be a dedicated limited-angle version of the CT-Fanbeam-180 experiment performed by Hu et al. \cite{hu_learning_2024}. Native-resolution experiments use 60 full-angle views on a $512\times512$ grid and compare FDK, CP-TV, Repo-Faithful PaDIS-VE-DPS and LION-Physical PaDIS-VE-DPS. We did not train a whole-image model at this resolution.

All quantitative ablations use the 20-view full-angle problem and the VE-DPS sampler. The patch-size, dataset-size and positional-encoding comparisons revisit investigations by Hu et al. \cite{hu_learning_2024}, while FDK vs noisy initialisation and geometric vs EDM noising-schedule ablations resolve paper--code discrepencies. Patch-size ablations cover the models trained and depicted in \autoref{tab:patch_size_schedules}. Dataset-size tests assesses the impact of using all available training set LIDC-IDRI slices for training over the default four slices per patient.

Unconditional generation presents four samples from each of four configurations. These are whole-image diffusion and the patch prior assembled using fixed stitching, fixed averaging or randomly shifted PaDIS. Sampling uses 300 geometric outer levels with one inner update per level. Separate tuning selected Langevin coefficient and noise scale pairs of $(0.75,0.75)$ for whole-image diffusion and $(0.8,0.85)$ for PaDIS. Both fixed patch methods use $(1,1)$.

Standard reconstruction jobs use seed 33 and the same first 25 test images. Fixed patch-averaging and patch-stitching rows, and $512\times512$ reconstructions use four samples because of their substantially greater cost and limited compute availability. Their metrics should therefore not be treated as equally precise estimates.

\section{Metrics}

Metric computation uses the held-out target in the common normalised-intensity domain with data range one. Reconstructions are evaluated without per-image rescaling, and clipping is applied only where specified by each method's configuration. This work reports mean absolute error (MAE), peak signal-to-noise ratio (PSNR), and structural similarity (SSIM) \cite{wang_image_2004}. For $N$ image pixels,
\begin{align}
\mathrm{MAE}(\hat{\mathbf{x}},\mathbf{x})
&=\frac{1}{N}\lVert\hat{\mathbf{x}}-\mathbf{x}\rVert_1,\\
\mathrm{PSNR}(\hat{\mathbf{x}},\mathbf{x})
&=10\log_{10}\left(\frac{1}{\mathrm{MSE}(\hat{\mathbf{x}},\mathbf{x})}\right),
\end{align}
where the peak value is one. SSIM measures local agreement in luminance, contrast and structure. Higher PSNR and SSIM and lower MAE indicate closer agreement, but none establishes that clinically important features have been preserved.

Measurement consistency is reported separately as the relative sinogram residual
\begin{equation}
\mathrm{RSR}(\hat{\mathbf{x}})
=
\frac{\lVert A\hat{\mathbf{x}}-\mathbf{y}\rVert_2}
{\lVert\mathbf{y}\rVert_2}.
\end{equation}
Low RSR shows that a reconstruction explains the simulated measurements. A low RSR does not, however, by itself imply similarity to the target because sparse- and limited-angle operators have large null and near-null spaces.

We report image-level means and asymmetric central 95\% confidence intervals, $\mathrm{mean}^{+u}_{-l}$ from 2000 deterministic bootstrap resamples using seed 33 \cite{efron_automatic_2020}, bootstrapping images rather than training runs to capture slice variation but not retraining uncertainty. For figures, we clip normalised images to $[0,1]$, and abbreviate ``Normalised Intensity'' to NI. HU panels use \autoref{eq:hounsfield_units}. We clip each row to the 15th and 95th percentiles of non-background ground-truth pixels, defined by normalised intensity above 0.02, and apply the same limits to every reconstruction and target in that row to prevent method-dependent contrast adjustment. Colour bars report these limits. The 50-mm scale bars use the configured 300-mm field of view rather than patient-native DICOM spacing, and we assess unconditional generations qualitatively because they have no paired target.
\section{Compute and Reproducibility}
For final training and reconstruction, we used a Google-hosted NVIDIA RTX PRO 6000 GPU, while fixed-validation tuning also used a local NVIDIA GTX 1070. Using seed 33 throughout, we developed the pipeline independently in LION, used the released repository only to check the Repo-Faithful track and selected checkpoints before testing. Training runs and stored artefacts are available through \href{https://wandb.ai/tjh200-university-of-cambridge/PaDIS-Reproduction?nw=nwusertjh200}{Weights \& Biases}. Reproduction scripts and instructions for their usage on Google Colab, Google Cloud Compute Engine, or a SLURM-enabled HPC cluster are provided \href{https://github.com/THartigan/MPhil-Project/tree/main/LION/scripts/paper_scripts/PaDIS-Reproduction}{here}.

% - Implemented in LION - Produced three implementation tracks: Paper-faithful
% implementation, repo-faithful matching setup, and one which uses correct operator
% normalisations to increase cross-problem reliability, as described in
% \ref{subsec:lion-physical-implementations}. - Created LION-Physical implementations for all
% experiments, but only paper-faithful and repo-faithful implementations for those
% which were actually described in sufficient detail for reproduction in those
% respective soruces. This is summarised in \autoref{tab:implemented-methods}.
